Static contract checking for Haskell

### 2011-08-22 paper.tex
paper fix

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage[pdftex]{graphicx}

@@

@@

\newcommand{\unr}{\texttt{UNR}}
\newcommand{\bad}{\texttt{BAD}}
\newcommand{\any}{\texttt{Any}}
\newcommand{\ok}{\texttt{Ok}}
\newcommand{\hprime}{\mathcal{H}'}
\newcommand{\cfc}{\texttt{CF}}
\newcommand{\cf}[1]{\texttt{CF}(#1)}
\newcommand{\weak}[1]{\mbox{$\$$weak}(#1)}
\renewcommand{\min}[1]{\mbox{$\$$min}(#1)}

\title{Static contract checking for Haskell}
\maketitle

@@

Finally, we define projections for $Cons$. It is not strictly
necessary, but it will be handy:
\begin{equation**}
\forall xs,y,ys.~sel_{1,Cons}(Cons(y,ys)) = xs \implies xs = y 
\end{equation**}
\begin{equation*}
\forall xs,y,ys.~sel_{2,Cons}(Cons(y,ys)) = xs \implies xs = ys 
\end{equation*}

@@

\begin{align*}
 \forall xs.~xs \neq \bad \land xs \neq Nil \notag \\
 \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to  \notag \\
 head(X) = \unr 
\end{algin*}

@@

\begin{itemize}
\item There's no $\lambda$-abstraction, because we can always lift
  them to toplevel declaration.
\item We do not allow nested case expressions, because once again, we
  can always lift them to the toplevel.
\item Until \label{ho} we will only consider full application of
  functions ($f(x,y)$), in order to remove clutter. Dealing with
  partial application is not hard but a bit cumbersome.
\end{itemize}

\begin{figure}[h]
  \centering
  \begin{array}{rclr}
    mod &:=& def_1,\dots,def_n &\\
    def &\in& \mbox{Definition} & \\
    def &:=& \texttt{data } T = K_1 \mid \cdots \mid K_n & \\
    &\mid& f \in c & \\
    &\mid& f~\vec{x}~=~e & \\
    &\mid& \hspace{-0.2cm}
       \begin{array}{rcl}
         f~\vec{x} &=& \texttt{case } e \texttt{ of } \\
         && K_1(\vec{x}) \to e_1 \mid \cdots \mid K_n(\vec{x}) \to e_n \\
       \end{array} & \\
  \end{array}

  \begin{array}{rclr}
    x,y,f,g,a,b & \in & \mbox{Variables} \\
    T &\in& \mbox{Type Constructors} \\
    K &\in& \mbox{Data Constructors} \\
  \end{array}

  \begin{array}{rclr}
    e &\in& \mbox{Expressions} & \\
    e &::=& x & \\
    &\mid& \bad & \\
    &\mid& e~e & \\
    &\mid& f(e,\dots,e) & \\
    &\mid& K(e,\dots,e) & \\
  \end{array}

@@

\begin{figure}[h]
  \label{cont-stx}
  \centering 
  \begin{array}{rclr}
  c &:=& x:c \to c\\
  &\mid& (c,c) \\
  &\mid& c \land c \\
  &\mid& c \lor c \\
  &\mid& \{ x \mid p \}\\
  &\mid& \cfc \\
  \end{array}
  \caption{Contract syntax}
\end{figure}

@@

\begin{figure}[h]
  \label{cont-smt}
  \centering
  \begin{array}{rcl}
    e \in \{x \mid p \} &\iff& e \mbox{ diverges or } p[e/x] \not \to^\star \{\bad,False\}\\
    e \in x:c_1 \to c_2 &\iff& \forall e_1 \in c_1, (e~e_1) \in c_2[e_1/x]\\
    e \in (c_1,c_2) &\iff& e \mbox{ diverges or }\\
    &&  (e \to^\star (e_1,e_2) \mbox{ and } e_1 \in c_1, e_2 \in c_2)\\
    e \in c_1 \land c_2 &\iff& e \in c_1 \mbox{ and } e \in c_2 \\
    e \in c_1 \lor c_2 &\iff& e \in c_1 \mbox{ or } e \in c_2 \\
    e \in \cfc &\iff& e \mbox{ is crash-free}
  \end{array}
  \caption{Semantics of contract satisfaction}
\end{figure}

@@

\begin{figure}
  \label{fol-stx}
  \centering
  \begin{array}{rcl}
    v,w,s,t &:=& x \mid K(t,\dots,t) \mid f(t,\dots,t) \mid app(t,t)\\
    && \mid \bad \mid \unr & \\
    \phi &:=& \forall x.\phi \mid \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t)\\
    && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi & \\
    \Phi &:=& \epsilon \mid \phi \mid \Phi \cup \Phi \\
  \end{array}
  \caption{First-order logic syntax}
\end{figure}

@@

\begin{figure*}
  \label{dtrans-case}
  $$ \dtrans{f(x_1,\dots,x_n) = \mbox{case } e \mbox{ of } [K_1(\vec{x_1}) = e_1,\dots,K_n(\vec{x_n}) = e_n]}$$
  \begin{center}
    \begin{array}{rccccccl}
      \forall \vec{a},y,\vec{x}_{1 \leq i \leq n}.~&\min{y} &\land& \etrans{e} = K_1(\vec{x_1}) &\land& f~\vec{a} = y &\to& y = \etrans{e_1}\\
                                               &&&&&& \cdots & \\
                                               &\min{y} &\land& \etrans{e} = K_n(\vec{x_n}) &\land& f~\vec{a} = y &\to& y = \etrans{e_n}\\
                                               &\min{y} &\land& \etrans{e} = \bad          &\land& f~\vec{a} = y &\to& y = \bad\\
                                               &\min{y} && \land                           &&      f~\vec{a} = y &\to& \min{a} \land (a=\bad \lor \bigvee_i a = \weak{K_i(\vec{x})} \lor y = \unr)
    \end{array}
    \end{center}
  \caption{$\dtrans{}$ -- Case definitions}
\end{figure*}

@@

\begin{itemize}
\item $(S_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i))$
  on its $k$-th component.
\item $(S_2)$ For each pair of constructors $K_i,K_j$, we state that they can
  never map to the same value.
\item $(S_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\leftarrow$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\rightarrow$: a value $t$ of type $T$ is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the k-th argument of $t$ is not crash-free, then the k-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

@@

\begin{figure*}
  \label{ttrans}
  $$ S_1 := \forall \vec{x},a.~(\min{a} \land K_i(\vec{x}) = a) \to \bigwedge_{1 \leq j \leq k} x_j = \weak{sel_{j,K_i}(a)}  \mid 1 \leq i \leq n$$
  $$ S_2 := \forall \vec{x},\vec{y},a,b.~(\min{a} \land \min{b} \land K_i(\vec{x}) = a \land K_j(\vec{y}) = b) \to a \neq b \mid 1 \leq i < j \leq n$$
  $$ S_3 := \forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to (\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{K_i(\vec{x})} \mid 1 \leq i \leq n$$
  $$ S_4 := \forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad \mid 1 \leq i \leq n$$
  \caption{$\ttrans{}$ -- Data type translation}
\end{figure*}

@@

\begin{figure*}
\begin{eqnarray*}
  \strans{e ::: \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = true \lor \etrans{b(e)} = \unr)\\
  \strans{e ::: x:c_1 \to c_2(x)} &=& \forall x.~\min{\etrans{f(x)}} \to (\strans{x :/: c_1} \lor  \strans{\etrans{e(x)} ::: c_2(x)})\\
  \strans{e ::: \cfc} &=& \cf{e}
\end{eqnarray*}

\begin{eqnarray*}
  \strans{e :/: \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} \neq true \land \etrans{b(e)} \land \unr)\\
  \strans{e :/: x:c_1 \to c_2(x)} &=& \exists x.~\min{\etrans{f(x)}} \to (\strans{x ::: c_1} \land  \strans{\etrans{e(x)} :/: c_2(x)})\\
  \strans{e :/: \cfc} &=& ~\cf{e}
\end{eqnarray*}

$$ \strans{e ::: c_1 && c_2} = \strans{e ::: c_1} \land \strans{e ::: c_2} $$
$$ \strans{e ::: c_1 || c_2} = \strans{e ::: c_1} \lor \strans{e ::: c_2} $$
$$ \strans{(a,b) ::: (c_1,c_2)}   = \strans{a ::: c_1} && \strans{b ::: c_2} $$

>>>>

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage[pdftex]{graphicx}

@@

\newtheorem{definition}{Definition}

@@

\newcommand{\unr}{\texttt{UNR}}
\newcommand{\bad}{\texttt{BAD}}
\newcommand{\any}{\texttt{Any}}
\newcommand{\ok}{\texttt{Ok}}
\newcommand{\hprime}{\mathcal{H}'}
\newcommand{\cfc}{\texttt{CF}}
\newcommand{\cf}[1]{\texttt{CF}(#1)}
\newcommand{\weak}[1]{\mbox{$\$$weak}(#1)}
\renewcommand{\min}[1]{\mbox{min}(#1)}

\title{Static contract checking for Haskell}
\authorinfo{Simon, Koen, Dimitrios, Charles-Pierre}
\maketitle

@@

Finally, we define projections for $Cons$. It is not strictly
necessary, but it will be handy:
\begin{equation*}
\forall xs,y,ys.~sel_{1,Cons}(Cons(y,ys)) = xs \implies xs = y 
\end{equation*}
\begin{equation*}
\forall xs,y,ys.~sel_{2,Cons}(Cons(y,ys)) = xs \implies xs = ys 
\end{equation*}

@@

\begin{align*}
 \forall xs.~xs \neq \bad \land xs \neq Nil \notag \\
 \land xs \neq Cons(sel_{1,Cons}(xs),sel_{2,Cons}(xs)) \to  \notag \\
 head(X) = \unr 
\end{align*}

@@

\begin{itemize}
\item There's no $\lambda$-abstraction, because we can always lift
  them to toplevel declaration.
\item We do not allow nested case expressions, because once again, we
  can always lift them to the toplevel.
\item Until \ref{ho} we will only consider full application of
  functions ($f(x,y)$), in order to remove clutter. Dealing with
  partial application is not hard but a bit cumbersome.
\end{itemize}

\begin{figure}[h]
  \centering
  \[  \begin{array}{rclr}
    mod &:=& def_1,\dots,def_n &\\
    def &\in& \mbox{Definition} & \\
    def &:=& \texttt{data } T = K_1 \mid \cdots \mid K_n & \\
    &\mid& f \in c & \\
    &\mid& f~\vec{x}~=~e & \\
    &\mid& \hspace{-0.2cm}
         \begin{array}{rcl}
         f~\vec{x} &=& \texttt{case } e \texttt{ of } \\
         && K_1(\vec{x}) \to e_1 \mid \cdots \mid K_n(\vec{x}) \to e_n \\
       \end{array} & \\
  \end{array} \]

  \[  \begin{array}{rclr}
    x,y,f,g,a,b & \in & \mbox{Variables} \\
    T &\in& \mbox{Type Constructors} \\
    K &\in& \mbox{Data Constructors} \\
  \end{array} \]

  \[  \begin{array}{rclr}
    e &\in& \mbox{Expressions} & \\
    e &::=& x & \\
    &\mid& \bad & \\
    &\mid& e~e & \\
    &\mid& f(e,\dots,e) & \\
    &\mid& K(e,\dots,e) & \\
  \end{array} \]

@@

\begin{figure}[h]
  \label{cont-stx}
  \centering 
  \[  \begin{array}{rclr}
  c &:=& x:c \to c\\
  &\mid& (c,c) \\
  &\mid& c \land c \\
  &\mid& c \lor c \\
  &\mid& \{ x \mid p \}\\
  &\mid& \cfc \\
  \end{array} \]
  \caption{Contract syntax}
\end{figure}

@@

\begin{figure}[h]
  \label{cont-smt}
  \centering
  \[  \begin{array}{rcl}
    e \in \{x \mid p \} &\iff& e \mbox{ diverges or } p[e/x] \not \to^\star \{\bad,False\}\\
    e \in x:c_1 \to c_2 &\iff& \forall e_1 \in c_1, (e~e_1) \in c_2[e_1/x]\\
    e \in (c_1,c_2) &\iff& e \mbox{ diverges or }\\
    &&  (e \to^\star (e_1,e_2) \mbox{ and } e_1 \in c_1, e_2 \in c_2)\\
    e \in c_1 \land c_2 &\iff& e \in c_1 \mbox{ and } e \in c_2 \\
    e \in c_1 \lor c_2 &\iff& e \in c_1 \mbox{ or } e \in c_2 \\
    e \in \cfc &\iff& e \mbox{ is crash-free}
  \end{array} \]
  \caption{Semantics of contract satisfaction}
\end{figure}

@@

\begin{figure}
  \label{fol-stx}
  \centering
  \[  \begin{array}{rcl}
    v,w,s,t &:=& x \mid K(t,\dots,t) \mid f(t,\dots,t) \mid app(t,t)\\
    && \mid \bad \mid \unr \\
    \phi &:=& \forall x.\phi \mid \lnot \phi \mid \phi \lor \phi \mid \top \mid \bot \mid t=t \mid \mbox{CF}(t)\\
    && \mid \phi \land \phi \mid \phi \to \phi \mid \phi \tot \phi  \\
    \Phi &:=& \epsilon \mid \phi \mid \Phi \cup \Phi \\
  \end{array} \]
  \caption{First-order logic syntax}
\end{figure}

@@

\begin{figure*}
  \label{dtrans-case}
  $$ \dtrans{f(x_1,\dots,x_n) = \mbox{case } e \mbox{ of } [K_1(\vec{x_1}) = e_1,\dots,K_n(\vec{x_n}) = e_n]}$$
  \begin{center}
    \[  \begin{array}{rccccccl}
      \forall \vec{a},y,\vec{x}_{1 \leq i \leq n}.~&\min{y} &\land& \etrans{e} = K_1(\vec{x_1}) &\land& f~\vec{a} = y &\to& y = \etrans{e_1}\\
                                               &&&&&& \cdots & \\
                                               &\min{y} &\land& \etrans{e} = K_n(\vec{x_n}) &\land& f~\vec{a} = y &\to& y = \etrans{e_n}\\
                                               &\min{y} &\land& \etrans{e} = \bad          &\land& f~\vec{a} = y &\to& y = \bad\\
                                               &\min{y} && \land                           &&      f~\vec{a} = y &\to& \min{e} \land (e=\bad \lor \bigvee_i e = \weak{K_i(\vec{x})} \lor y = \unr)
    \end{array} \]
    \end{center}
  \caption{$\dtrans{}$ -- Case definitions}
\end{figure*}

@@

\begin{itemize}
\item $(S_1)$ For each $K_i$ of arity $a_i$ we introduce selectors
  $sel_{k,K_i}$, which are the projection of $K_i(x_1,\dots,x_{n_i})$
  on its $k$-th component.
\item $(S_2)$ For each pair of constructors $K_i,K_j$, we state that they can
  never map to the same value.
\item $(S_3)$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\leftarrow$: if we pack crash-free values in a data
    constructor, the resulting value is crash-free.
  \item $\rightarrow$: a value $t$ of type $T$ is crash-free implies
    that every value packed in it is crash-free. Recall that one can
    define projection on any argument of a value of type $t$. So if
    the k-th argument of $t$ is not crash-free, then the k-th
    projection is a crash-free context that throws an expression that
    is not crash-free.

@@

\begin{figure*}
  \label{ttrans}
  $$ S_1 := \forall \vec{x},a.~(\min{a} \land K_i(\vec{x}) = a) \to \bigwedge_{1 \leq j \leq k} x_j = \weak{sel_{j,K_i}(a)}  \mid 1 \leq i \leq n$$
  $$ S_2 := \forall \vec{x},\vec{y},a.~ \lnot (\min{a} \land K_i(\vec{x}) = a \land K_j(\vec{x}) = a) \mid 1 \leq i < j \leq n$$
  $$ S_3 := \forall \vec{x},a.~\min{a} \land a = K_i(\vec{x}) \to ((\bigwedge_{1 \leq j \leq k} \cf{x_j}) \tot \cf{K_i(\vec{x})}) \mid 1 \leq i \leq n$$
  $$ S_4 := \forall \vec{x},a.~(\min{a} \land a = K_i(\vec{x})) \to a \neq \unr \land a \neq \bad \mid 1 \leq i \leq n$$
  \caption{$\ttrans{}$ -- Data type translation}
\end{figure*}

@@

\begin{figure*}
\begin{eqnarray*}
  \strans{e \in \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = true \lor \etrans{b(e)} = \unr)\\
  \strans{e \in x:c_1 \to c_2(x)} &=& \forall x.~\min{\etrans{f(x)}} \to (\strans{x \not \in c_1} \lor  \strans{\etrans{e(x)} \in c_2(x)})\\
  \strans{e \in \cfc} &=& \cf{e}
\end{eqnarray*}

\begin{eqnarray*}
  \strans{e \not \in \{x \mid b(x) \}} &=&  \min{\etrans{b(e)}} \land (\etrans{b(e)} = false \lor \etrans{b(e)} = \bad)\\
  \strans{e \not \in x:c_1 \to c_2(x)} &=& \exists x.~ \strans{x \in c_1} \land  \strans{\etrans{e(x)} \not \in c_2(x)}\\
  \strans{e \not \in \cfc} &=& \lnot \cf{e}
\end{eqnarray*}

\begin{center}
\[  \begin{array}{rcl}
 \strans{e \in c_1 \&\& c_2} &=& \strans{e \in c_1} \land \strans{e \in c_2} \\
 \strans{e \not \in c_1 \&\& c_2} &=& \strans{e \not \in c_1} \lor \strans{e \not \in c_2}\\
 \strans{e \in c_1 || c_2} &=& \strans{e \in c_1} \lor \strans{e \in c_2} \\
  \strans{e \not \in c_1 || c_2} &=& \strans{e \not \in c_1} \land \strans{e \not \in c_2}\\
 \strans{(a,b) \in (c_1,c_2)}   &=& \strans{a \in c_1} \land \strans{b \in c_2} \\
 \strans{(a,b) \not \in (c_1,c_2)}   &=& \strans{a \not \in c_1} \lor \strans{b \not \in c_2} \\
\end{array} \]
\end{center}
